---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse) 
library(here) 
library(ggplot2) 

#Incidence is across 2013 through 2017 and Mortality is across 2014 through 2018 
simd <- read_csv(here("clean_data/cancer_by_simd_scotland.csv")) 

incidences <- read_csv(here("clean_data/cancer_incidences_in_scotland.csv")) 

mortality <- read_csv(here("clean_data/cancer_mortality_in_scotland.csv")) 

#These diagnosed cases are only for malignant brain tumours (CID 71) 
england_newly_diagnosed_cases <- read_csv(here("clean_data/newly_diagnosed_england.csv")) 

post_surgery_experience_2018 <- readxl::read_xlsx(here("clean_data/raw_data_post_surgery_experience_2018.xlsx")) 

tumour_grades_2018 <- read_csv(here("clean_data/raw_data_tumour_grades.csv")) 

prevalence <- read_csv(here("clean_data/scotland_cancer_prevalence_2017.csv"))  

updated_cancer_incidences <- read_csv(here("clean_data/more_recent_cancer_incidences.csv")) 

survival_data <- read_csv(here("raw_data_cancer_survival.csv")) 

survival_leukaemia_clean <- read_csv(here("survival_leukaemia_clean.csv")) 
```

Themes for the graphs. 

```{r}
title.theme <- element_text(colour = "brown3", size = 14, face = "bold")
subtitle.theme <- element_text(colour = "black", size = 12, face = "bold") 
x.axis.theme <- element_text(face = "bold.italic") 
```


#Comparing England and Scotland 

Plotting the number of cases of malignant brain cancer in England from 2008 to 
2017 to get an understanding of the trend. 

```{r}
incidences_joined <- england_newly_diagnosed_cases %>% 
  mutate(hb_label = "England") %>% 
  mutate(sex_label = gender) %>% 
  select(-gender) %>% 
  mutate(incidence_no_and_rates = raw_cases) %>% 
  full_join(scotland_incidences_all_ages) %>% 
  mutate(hb_label = recode(hb_label, "SCOTLAND" = "Scotland")) 

incidences_joined %>% 
  group_by(year) %>% 
  summarise(total = incidence_no_and_rates) 

2654 + 1914 

267 + 199 
```


```{r} 
#This data is only for malignant brain tumours. For this reason the comparison graph below for Scotland is also only showing recorded malignant brain tumours. 
incidences_joined %>% 
  group_by(year, sex_label, hb_label) %>% 
  summarise(total = sum(incidence_no_and_rates)) %>% 
  filter(hb_label == "Scotland" | hb_label == "England") %>% 
  ggplot(aes(year, total, fill = sex_label)) + 
  geom_col() + 
  scale_x_continuous(breaks = seq(2008, 2017, by = 1)) + 
  labs(title = "Diagnosed Cases of malignant brain cancer in England and Scotland", 
       subtitle = "2008-2017") + 
  xlab("Year") + 
  ylab("Number of Cases") + 
  theme(
  plot.title = title.theme, 
  plot.subtitle = subtitle.theme, 
  axis.title = x.axis.theme) + 
  scale_fill_manual(values = c("palevioletred4", "cadetblue3")) + 
  guides(fill = guide_legend(title = "Sex")) + 
  facet_grid(~ hb_label) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 0.9)) 
```

```{r}
ggsave("diagnosed_cases_in_eng.png", plot = last_plot())  
```


Plotting the same graph but for Scottish data to see how the two compare. 

```{r}
scotland_incidences_all_ages <- incidences %>% 
  filter(hb_label == "SCOTLAND") %>% 
  filter(age_label == "Number of registrations: All Ages") %>% 
  filter(site_label == "Malignant brain cancer (incl pituitary gland, craniopharyngeal duct and pineal gland)") %>% 
  filter(sex_label == "Males" | sex_label == "Females") %>% 
  filter(year > 2007) 

scotland_incidences_all_ages %>% 
  ggplot(aes(year, incidence_no_and_rates, fill = sex_label)) + 
  geom_col() + 
  scale_x_continuous(breaks = seq(2008, 2017, by = 1)) + 
  labs(title = "Diagnosed Cases of malignant brain cancer in Scotland", 
       subtitle = "2008-2017") + 
  xlab("Year") + 
  ylab("Number of Cases") + 
  theme(
  plot.title = title.theme, 
  plot.subtitle = subtitle.theme, 
  axis.title = x.axis.theme) + 
  scale_fill_manual(values = c("palevioletred4", "cadetblue3")) + 
  guides(fill = guide_legend(title = "Sex"))  
```

```{r}
ggsave("diagnosed_cases_in_scotland.png", plot = last_plot())  
```


# Daily Cases 

https://beta.isdscotland.org/find-publications-and-data/conditions-and-diseases/cancer/cancer-incidence-in-scotland/28-april-2020/ 

Identifying the number of cases diagnosed each day from 1994 to 2018. This 
includes all three brain cancer types specified in the data (Malignant 
Brain Cancer, Malignant Brain Cancer including the pituitary gland, 
craniopharyngeal duct and pineal gland and Non Malignant cancer including the 
pituitary gland, craniopharyngeal duct and pineal gland). 

I decided to find the mean number of incidences of non-malignant brain tumours 
in order to impute the missing values with it. According to Wu, Khorshidi, 
Aickelin, Edib and Peate (2019), simple mean mutation is a better method of 
handling missing values than deletion when dealing with breast cancer 
data. https://link.springer.com/article/10.1007/s13755-019-0082-4 

```{r}
updated_cancer_incidences %>% 
  filter(hb_label == "SCOTLAND") %>% 
  filter(age_label == "Number of registrations: All Ages") %>% 
  filter(site_label == "Non-malignant brain cancer (incl pituitary gland, craniopharyngeal duct and pineal gland)") %>% 
  filter(sex_label == "All Persons") %>% 
  drop_na() %>% 
  summarise(mean_incidence = mean(incidence_no_and_rates)) 
```

```{r} 
#Filtering out malignant brain cancer and total cancer cases because I am trying to ascertain if the malignant brain cancer including pituitary gland etc and non-malignant brain cancer including the pituitary gland etc combine to make the same total as the all brain and CNS tumours incidence column.  
scotland_incidences_all_types <- updated_cancer_incidences %>% 
  filter(hb_label == "SCOTLAND") %>% 
  filter(age_label == "Number of registrations: All Ages") %>% 
  filter(sex_label == "All Persons") %>% 
  filter(site_label != "Malignant brain cancer") %>% 
  filter(site_label != "All brain and CNS tumours (malignant and non-malignant)") %>% 
  mutate(incidence_no_and_rates = coalesce(incidence_no_and_rates, 534, na.rm = TRUE))  
```

The total number of cases of malignant and non-malignant brain tumours including 
those in the pituitary gland, craniopharyngeal duct and pineal gland is the 
same as the total number of tumours found in the brain and central nervous 
system data. However, it was necessary to identify the total number of cases by 
hand in order to impute the missing values from 1994 to 1999 in the 
non-malignant cancer data with the mean. The data for total tumours for those
years was missing and so had to be calculated as above. 

```{r} 
scotland_incidences_all_types %>% 
  group_by(year) %>% 
  summarise(total_cases = sum(incidence_no_and_rates)) %>% 
  mutate(avg_diagnosed_per_day = total_cases / 365) %>% 
  ggplot(aes(year, avg_diagnosed_per_day)) + 
  geom_point() + 
  geom_smooth(se = FALSE, color = "brown3") + 
  scale_x_continuous(breaks = seq(1994, 2018, by = 1)) + 
  labs(title = "Estimated Daily Diagnosis in Scotland from 1993 to 2017", 
       subtitle = "Malignant and non-malignant brain tumours") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 0.9)) + 
  xlab("Year of Diagnosis") + 
  ylab("Estimated New Cases Each Day") + 
  theme(
  plot.title = title.theme, 
  plot.subtitle = subtitle.theme, 
  axis.title = x.axis.theme) 
```

```{r}
scotland_incidences_all_types %>% 
  group_by(year) %>% 
  summarise(total_cases = sum(incidence_no_and_rates)) %>% 
  ggplot(aes(year, total_cases)) + 
  geom_point() + 
  geom_smooth(se = FALSE, color = "brown3") + 
  scale_x_continuous(breaks = seq(1994, 2018, by = 1)) + 
  labs(title = "Malignant and Non-Malignant Tumour Diagnosis in Scotland", 
       subtitle = "1993-2017") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 0.9)) + 
  xlab("Year of Diagnosis") + 
  ylab("Incidences Each Year") + 
  theme(
  plot.title = title.theme, 
  plot.subtitle = subtitle.theme, 
  axis.title = x.axis.theme) 
```

```{r}
ggsave("diagnosis_of_all_tumours.png", plot = last_plot()) 
```


# Percentage of the Scottish population 

As of mid 2018, the Scottish population was 5,438,100 (National Records of 
Scotland, https://www.nrscotland.gov.uk/files/statistics/rgar/2018/rgar18.pdf). 
The number of cases diagnosed in 2018 was 966. 

The percentage of the Scottish population in 2018 diagnosed with a brain tumour was 0.018%. In 2000, the percentage of the population diagnosed with a brain tumour was 0.016%. The difference is pretty minor, but see above for a closer look at estimated daily cases of brain tumours from 1993 to 1997.  

```{r}
updated_cancer_incidences %>% 
  filter(hb_label == "SCOTLAND") %>% 
  filter(age_label == "Number of registrations: All Ages") %>% 
  filter(sex_label == "All Persons") %>% 
  filter(site_label == "All brain and CNS tumours (malignant and non-malignant)") %>% 
  filter(year == 2018) 

updated_cancer_incidences %>% 
  filter(hb_label == "SCOTLAND") %>% 
  filter(age_label == "Number of registrations: All Ages") %>% 
  filter(sex_label == "All Persons") %>% 
  filter(site_label == "All brain and CNS tumours (malignant and non-malignant)") %>% 
  filter(year == 2000) 

966 / 5438100 * 100 

807 / 5067000 * 100 
```

```{r}
updated_cancer_incidences %>% 
  filter(site_label == "All brain and CNS tumours (malignant and non-malignant)") %>% 
  filter(sex_label == "All Persons") %>% 
  filter(age_label == "Crude Rate (Incidence)") %>% 
  filter(hb_label != "SCOTLAND") %>% 
  filter(hb_label != "North of Scotland") %>% 
  filter(hb_label != "South East of Scotland") %>% 
  filter(hb_label != "West of Scotland") %>% 
  filter(year == 2017) %>% 
  ggplot(aes(hb_label, incidence_no_and_rates)) + 
  geom_col() + 
  labs(title = "Crude Rate of Brain and CNS Tumour Incidences", subtitle = "2017") + 
  xlab("Health Board") + 
  ylab("Crude Rate of Diagnosis") + 
  theme(axis.text.x = element_text(angle = 35, hjust = 0.9)) + 
  theme(
  plot.title = title.theme, 
  plot.subtitle = subtitle.theme, 
  axis.title = x.axis.theme) 
```

```{r}
ggsave("crude_rate_by_area_in_2017.png", last_plot()) 
```


```{r}
updated_cancer_incidences %>% 
  filter(sex_label == "All Persons") %>% 
  filter(site_label == "All brain and CNS tumours (malignant and non-malignant)") %>% 
  filter(hb_label != "SCOTLAND") %>% 
  filter(hb_label != "North of Scotland") %>% 
  filter(hb_label != "South East of Scotland") %>% 
  filter(hb_label != "West of Scotland") %>% 
  filter(age_label == "Crude Rate (Incidence)") %>% 
  filter(year == 2010) %>% 
  ggplot(aes(hb_label, incidence_no_and_rates)) + 
  geom_col() + 
  theme(axis.text.x = element_text(angle = 35, hjust = 0.9)) + 
  labs(title = "Crude Rate of Brain and CNS Tumour Incidence", subtitle = "2010") + 
  xlab("Health Board") + 
  ylab("Crude Rate") + 
  theme(
  plot.title = title.theme, 
  plot.subtitle = subtitle.theme, 
  axis.title = x.axis.theme) 
```

```{r}
ggsave("crude_rate_by_area_in_2010.png", last_plot()) 
```

```{r}
updated_cancer_incidences %>% 
  filter(site_label == "All brain and CNS tumours (malignant and non-malignant)") %>% 
  filter(sex_label == "All Persons") %>% 
  filter(age_label == "Crude Rate (Incidence)") %>% 
  filter(hb_label != "SCOTLAND") %>% 
  filter(hb_label != "North of Scotland") %>% 
  filter(hb_label != "South East of Scotland") %>% 
  filter(hb_label != "West of Scotland") %>% 
  filter(year == 2000) %>% 
  ggplot(aes(hb_label, incidence_no_and_rates)) + 
  geom_col() + 
  labs(title = "Crude Rate of Brain and CNS Tumour Incidences", subtitle = "2000") + 
  xlab("Health Board") + 
  ylab("Crude Rate of Diagnosis") + 
  theme(axis.text.x = element_text(angle = 35, hjust = 0.9)) + 
  theme(
  plot.title = title.theme, 
  plot.subtitle = subtitle.theme, 
  axis.title = x.axis.theme) 
```

```{r}
ggsave("crude_rate_by_area_in_2000.png", last_plot()) 
```


# Age Demographic Cancer Incidence 

Identifying the percentage of people diagnosed in 2017 with brain cancer who 
are over 54. 

```{r}
updated_cancer_incidences %>% 
  filter(year == 2017) %>% 
  filter(site_label == "All brain and CNS tumours (malignant and non-malignant)") %>% 
  filter(hb_label == "SCOTLAND") %>% 
  filter(sex_label == "All Persons") %>% 
  filter(age_label == "Number of registrations: Under 5" | age_label ==  "Number of registrations: 5.." | age_label == "Number of registrations: 10-14" | age_label ==  "Number of registrations: 15-19" | age_label == "Number of registrations: 20-24" | age_label ==  "Number of registrations: 25-29" | age_label == "Number of registrations: 30-34" | age_label ==  "Number of registrations: 35-39"| age_label ==  "Number of registrations: 40-44" | age_label ==  "Number of registrations: 45-49" | age_label ==  "Number of registrations: 50-54" | age_label ==  "Number of registrations: 55-59" | age_label ==  "Number of registrations: 60-64" | age_label == "Number of registrations: 65-69" | age_label ==  "Number of registrations: 70-74" | age_label == "Number of registrations: 75-79" | age_label == "Number of registrations: 80x4" | age_label ==  "Number of registrations: 85x9" | age_label ==  "Number of registrations: 90+") %>% 
  summarise(total_incidences_in_2017 = sum(incidence_no_and_rates)) 

updated_cancer_incidences %>% 
  filter(year == 2017) %>% 
  filter(site_label == "All brain and CNS tumours (malignant and non-malignant)") %>% 
  filter(hb_label == "SCOTLAND") %>% 
  filter(sex_label == "All Persons") %>% 
  filter(age_label ==  "Number of registrations: 55-59" | age_label ==  "Number of registrations: 60-64" | age_label == "Number of registrations: 65-69" | age_label ==  "Number of registrations: 70-74" | age_label == "Number of registrations: 75-79" | age_label == "Number of registrations: 80x4" | age_label ==  "Number of registrations: 85x9" | age_label ==  "Number of registrations: 90+") %>% 
  summarise(total_incidences_in_2017 = sum(incidence_no_and_rates)) 

740 / 1077 * 100 
```


# Survival 

Net survival is an estimate of the probability that the average patient will be
alive at a given time after diagnosis, after controlling for other causes
of death. 

£15038718 spent on Brain cancer research in 2018/2019 (6.55%)
£35439980 spent on Leukaemia research in 2018/2019 (15.43%) 

In 2012/2013, only £6 million was spent on Brain cancer research. The allocated
budget for brain tumour research is growing, but there is still a bridge to gap 
between this region and research into other cancers. This may be an important 
factor in reducing the preventable deaths that still occur for those
diagnosed with a brain tumour. 

```{r}
joined <- full_join(survival_data, survival_leukaemia) 
```


```{r}
#Net survival of males aged from 15 to 99 at 1 and 5 years after the diagnosis. 
joined %>% 
  select(cancer_site_grouping, calendar_period_of_diagnosis, sex, age_group, years_since_diagnosis, 
         net_survival_percent) %>% 
  mutate(net_survival_percent = as.double(net_survival_percent)) %>%  
  mutate(years_since_diagnosis = as.factor(years_since_diagnosis)) %>% 
  filter(sex == "Male") %>% 
  filter(calendar_period_of_diagnosis == "2013-2017" | 
            calendar_period_of_diagnosis == "2008-2012") %>% 
  filter(age_group == "15-99") %>% 
  filter(years_since_diagnosis != 10) %>% 
  ggplot(aes(years_since_diagnosis, net_survival_percent, fill = cancer_site_grouping)) + 
  geom_bar(stat= "identity", position = "dodge") + 
  facet_wrap(~ calendar_period_of_diagnosis) + 
  labs(title = "Brain and CNS Tumour and Leukaemia Net Survival (Male)", 
       subtitle = "Estimated probablity of survival for 1 year and 5 years after diagnosis") + 
  xlab("Years Since Diagnosis") + 
  ylab("Net Survival (%)") + 
  theme(
  plot.title = title.theme, 
  plot.subtitle = subtitle.theme, 
  axis.title = x.axis.theme) + 
  guides(fill = guide_legend(title = "Cancer Site")) + 
  scale_y_continuous(breaks = seq(0, 100, by = 5)) + 
  scale_fill_manual(values = c("palevioletred4", "cadetblue3")) 
```

```{r}
ggsave("survival_male.png", last_plot()) 
```


```{r}
#Net survival of females aged from 15 to 99 at 1 and 5 years after the diagnosis. 
joined %>% 
  select(cancer_site_grouping, calendar_period_of_diagnosis, sex, age_group, years_since_diagnosis, net_survival_percent) %>% 
  mutate(net_survival_percent = as.double(net_survival_percent)) %>%  
  mutate(years_since_diagnosis = as.factor(years_since_diagnosis)) %>% 
  filter(sex == "Female") %>% 
  filter(calendar_period_of_diagnosis == "2013-2017" | 
            calendar_period_of_diagnosis == "2008-2012") %>% 
  filter(age_group == "15-99") %>% 
  filter(years_since_diagnosis != 10) %>% 
  ggplot(aes(years_since_diagnosis, net_survival_percent, fill = cancer_site_grouping)) + 
  geom_bar(stat= "identity", position = "dodge") + 
  facet_wrap(~ calendar_period_of_diagnosis) + 
  labs(title = "Brain and CNS Tumour Net Survival (Female)", 
       subtitle = "Estimated probablity of survival for 1 year and 5 years after diagnosis") + 
  xlab("Years Since Diagnosis") + 
  ylab("Net Survival (%)") + 
  scale_y_continuous(breaks = seq(0, 100, by = 5)) + 
  theme(
  plot.title = title.theme, 
  plot.subtitle = subtitle.theme, 
  axis.title = x.axis.theme) + 
  guides(fill = guide_legend(title = "Cancer Site")) + 
  scale_fill_manual(values = c("palevioletred4", "cadetblue3")) 
```

```{r}
ggsave("survival_female.png", last_plot()) 
```


There was missing data in the 10 year survival rates of brain tumours, 
however there is data for the number of deaths within 10 years of a diagnosis 
of brain cancer. The number of deaths decreased slightly in the 2013-2017 
grouping when compared to the 2008-2012 grouping, suggesting more people 
are surviving brain cancer for longer. Over this time period, 78% of deaths 
were in those over the age of 54, suggesting death may be more likely to occur
in those over 54. However it is important to note that those aged over 54 
made up 68% of the recorded cases in 2017, so the higher number of deaths 
seen in patients over 54 may be in part to their large representation in 
the data. 

```{r}
survival_data %>% 
  select(sex, age_group, years_since_diagnosis, deaths_within_10_years, calendar_period_of_diagnosis) %>% 
  filter(years_since_diagnosis == 1) %>% 
  filter(calendar_period_of_diagnosis == "2008-2012" 
         | calendar_period_of_diagnosis ==  "2013-2017") %>% 
  filter(age_group == "15-44" | age_group == "45-54" | age_group == "55-64" | 
          age_group == "65-74" | age_group == "75-99") %>% 
  summarise(total_ages = sum(deaths_within_10_years)) 

survival_data %>% 
  select(sex, age_group, years_since_diagnosis, deaths_within_10_years, calendar_period_of_diagnosis) %>% 
  filter(years_since_diagnosis == 1) %>% 
  filter(calendar_period_of_diagnosis == "2008-2012" 
         | calendar_period_of_diagnosis ==  "2013-2017") %>% 
  filter(age_group == "55-64" | 
          age_group == "65-74" | age_group == "75-99") %>% 
  summarise(total_55_plus = sum(deaths_within_10_years)) 

2720 / 3486	* 100 
```

```{r}
#Deaths across calendar periods 
survival_data %>% 
  select(sex, age_group, years_since_diagnosis, deaths_within_10_years, calendar_period_of_diagnosis) %>% 
  filter(years_since_diagnosis == 1) %>% 
  filter(calendar_period_of_diagnosis == "2008-2012" 
         | calendar_period_of_diagnosis ==  "2013-2017") %>% 
  filter(age_group == "15-44" | age_group == "45-54" | age_group == "55-64" | 
          age_group == "65-74" | age_group == "75-99") %>% 
  group_by(calendar_period_of_diagnosis) %>% 
  summarise(total = sum(deaths_within_10_years))
```


```{r}
#Deaths within 10 years 
survival_data %>% 
  select(sex, age_group, years_since_diagnosis, deaths_within_10_years, calendar_period_of_diagnosis) %>% 
  filter(years_since_diagnosis == 1) %>% 
  filter(calendar_period_of_diagnosis == "2008-2012" 
         | calendar_period_of_diagnosis ==  "2013-2017") %>% 
  filter(age_group == "15-44" | age_group == "45-54" | age_group == "55-64" | 
          age_group == "65-74" | age_group == "75-99") %>% 
  ggplot(aes(calendar_period_of_diagnosis, deaths_within_10_years, fill = age_group)) + 
  geom_col() + 
  labs(title = "Deaths within 10 years of a diagnosis of Brain and CNS cancer", 
       subtitle = "2008-2017") + 
  xlab("Calendar Period of Diagnosis") + 
  ylab("Number of Deaths within 10 years") + 
  theme(
  plot.title = title.theme, 
  plot.subtitle = subtitle.theme, 
  axis.title = x.axis.theme) + 
  guides(fill = guide_legend(title = "Age Group")) + 
  scale_fill_manual(values = c("slategray1", "slategray2", "slategray3", 
                                "slategray4", "gray29")) 
```

```{r}
ggsave("years_mortality.png", last_plot())  
```


# Mortality 

Mortality appears to vary by region each year. 

```{r}
mortality %>% 
  filter(site_label == "Malignant brain cancer (incl pituitary gland, craniopharyngeal duct and pineal gland)") %>% 
  filter(sex_label == "All Persons") %>% 
  filter(age_label == "Crude Rate (Mortality)") %>% 
  filter(hb_label != "SCOTLAND") %>% 
  filter(hb_label != "North of Scotland") %>% 
  filter(hb_label != "South East of Scotland") %>% 
  filter(hb_label != "West of Scotland") %>% 
  filter(year == 2017) %>% 
  ggplot(aes(hb_label, incidence_no_and_rates)) + 
  geom_col() + 
  labs(title = "Crude Rate of Mortality from Brain and CNS Cancer by Health Board", 
       subtitle = "2017") + 
  xlab("Health Board") + 
  ylab("Crude Rate of Death") + 
  theme(axis.text.x = element_text(angle = 35, hjust = 0.9)) + 
  theme(
  plot.title = title.theme, 
  plot.subtitle = subtitle.theme, 
  axis.title = x.axis.theme) 
```

```{r}
mortality %>% 
  filter(site_label == "Malignant brain cancer (incl pituitary gland, craniopharyngeal duct and pineal gland)") %>% 
  filter(sex_label == "All Persons") %>% 
  filter(age_label == "Crude Rate (Mortality)") %>% 
  filter(hb_label != "SCOTLAND") %>% 
  filter(hb_label != "North of Scotland") %>% 
  filter(hb_label != "South East of Scotland") %>% 
  filter(hb_label != "West of Scotland") %>% 
  filter(year == 2010) %>% 
  ggplot(aes(hb_label, incidence_no_and_rates)) + 
  geom_col() + 
  labs(title = "Crude Rate of Mortality from Brain and CNS Cancer by Health Board", 
       subtitle = "2010") + 
  xlab("Health Board") + 
  ylab("Crude Rate of Death") + 
  theme(axis.text.x = element_text(angle = 35, hjust = 0.9)) + 
  theme(
  plot.title = title.theme, 
  plot.subtitle = subtitle.theme, 
  axis.title = x.axis.theme) 
```

```{r}
mortality %>% 
  filter(site_label == "Malignant brain cancer (incl pituitary gland, craniopharyngeal duct and pineal gland)") %>% 
  filter(sex_label == "All Persons") %>% 
  filter(age_label == "Crude Rate (Mortality)") %>% 
  filter(hb_label != "SCOTLAND") %>% 
  filter(hb_label != "North of Scotland") %>% 
  filter(hb_label != "South East of Scotland") %>% 
  filter(hb_label != "West of Scotland") %>% 
  filter(year == 1993) %>% 
  ggplot(aes(hb_label, incidence_no_and_rates)) + 
  geom_col() + 
  labs(title = "Crude Rate of Mortality from Brain and CNS Cancer by Health Board", 
       subtitle = "1993") + 
  xlab("Health Board") + 
  ylab("Crude Rate of Death") + 
  theme(axis.text.x = element_text(angle = 35, hjust = 0.9)) + 
  theme(
  plot.title = title.theme, 
  plot.subtitle = subtitle.theme, 
  axis.title = x.axis.theme) 
```

Over the period of 2014 to 2018, the risk of death per 100,000 people at risk 
varied across SIMD quintiles. For those in the second least deprived quintile 
the mortality rate was highest at 8.3, however the second most deprived 
quintile is not far behind at 8.2. The lowest mortality rate was in the 
most deprived quintile, with a rate of 6.6. 

```{r}
simd %>% 
  ggplot(aes(simd_2016_deprivation_quintile, easr_1)) + 
  geom_col() + 
  xlab("SIMD Deprivation Quintile") + 
  ylab("Rate of Mortality") + 
  labs(title = "Mortality Rates by SIMD Deprivation Quintile", subtitle = 
         "2014-2018") + 
  theme(
  plot.title = title.theme, 
  plot.subtitle = subtitle.theme, 
  axis.title = x.axis.theme) + 
   scale_fill_manual(values = c("slategray1", "slategray2", "slategray3", 
                                "slategray4", "gray29")) 
```

```{r}
ggsave("deprivation_mortality.png", last_plot())  
```


# Patient Experiences 

In 2018, the Scottish Adult Neuro-Oncology Network published an audit with some 
statistics regarding patient experience. The target for 30 day mortality rate 
is below 5%. The only region to be over the target was Aberdeen at 9.8%. After 
surgery, imaging should be done within 3 days to measure surgical performance 
and determine necessary treatment. The target is 90%, however both Aberdeen 
and Dundee were under this 90% target. 

```{r}
library(reshape2)

post_surgery_experience_2018 %>% 
  pivot_longer(cols = c(proportion_of_patients_die_within_30_days_of_surgery, proportion_of_patients_imaging_within_72_hours), names_to = "surgery_aftermaths", 
               values_to = "proportion") %>% 
  mutate(surgery_aftermaths = recode(surgery_aftermaths, "proportion_of_patients_die_within_30_days_of_surgery" = "Patients dying within 30 days", 
"proportion_of_patients_imaging_within_72_hours" = "Patients recieving brain imaging within 3 days")) %>% 
  ggplot(aes(location, proportion, fill = surgery_aftermaths)) + 
  geom_bar(stat= "identity", position = "dodge") + 
  guides(fill = guide_legend(title = "Surgery Aftermath")) + 
  labs(title = "Surgery Aftermath by Location", 
       subtitle = "2018") + 
  xlab("Region") + 
  ylab("Percentage of Patients") + 
  theme(
  plot.title = title.theme, 
  plot.subtitle = subtitle.theme, 
  axis.title = x.axis.theme) + 
  scale_fill_manual(values = c("palevioletred4", "cadetblue3")) 
```

```{r}
ggsave("surgery_aftermath.png", last_plot()) 
```

